---
title: "p8105_hw3_geg2145"
author: "Gustavo Garcia-Franceschini"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(p8105.datasets)
```

# Problem 1

```{r p1_data}
data("instacart")
head(instacart, 4) %>%
  knitr::kable()
```

**The dataset has `r nrow(instacart)` rows and `r ncol(instacart)` variables, where every row is a product ordered. Every product ordered has an `order_id` (it is possible for multiple products to be bought as part of the same order) a `product_name` and an `aisle` name. We can also see the day of the week and hour the order was processed (`order_dow` and `order_hour_of_day`, respectively).**

How many aisles are there, and which aisles are the most items ordered from?

```{r p1_q1}
instacart %>%
  group_by(aisle) %>%
  summarize(items_ordered = n()) %>%
  arrange(desc(items_ordered)) %>% 
  head(4)
```

**There are `r length(unique(pull(instacart, aisle)))` different aisles, with most of the orders being in the "fresh vegetables", "fresh fruits", "packaged vegetables fruits" and "yogurt" aisles.**

Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r p1_plot}
instacart %>%
  group_by(aisle) %>%
  summarize(items_ordered = n()) %>%
  arrange(desc(items_ordered)) %>%
  filter(items_ordered > 10000) %>%
  mutate(aisle = fct_reorder(aisle, items_ordered)) %>%
  ggplot() + geom_point(aes(x = aisle, y = items_ordered)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title= "Best selling aisles sell food and perishables",
       x="Aisle", "Items Ordered")
```

**We can also see here how most of the orders are from the "fresh vegetables", "fresh fruits", "packaged vegetables fruits" and "yogurt" aisles. We can also see "butter", "oils and vinegars" and "dry pasta" are the aisles least bought from. We also see how most of the aisles (about two thirds) have less than 20,000 orders, and then the number of orders increases exponentially. The aisles with the most orders usually hold food that is consumed often (like water) and/or goes bad easily (like milk).**

Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r p1_table1}
instacart %>%
  filter(aisle == "baking ingredients" | 
         aisle == "dog food care" |
         aisle == "packaged vegetables fruits") %>%
  group_by(aisle, product_name) %>%
  summarize(times_ordered = n()) %>%
  mutate(rank = min_rank(desc(times_ordered))) %>%
  filter(rank <= 3) %>%
  arrange(aisle, rank) %>%
  knitr::kable()
```
**The best selling baking ingredient is light brown sugar, while the best selling packaged vegetables and fruits is organic baby spinach. The three highest selling dog food care times have similar order counts.**

    
Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r p1_table2}
instacart %>%
  filter(product_name == "Pink Lady Apples" | 
         product_name == "Coffee Ice Cream") %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day)) %>%
  pivot_wider(names_from = order_dow, 
              values_from = mean_hour) %>%
  knitr::kable()
```

**Coffee Ice Cream is bought in the afternoon (around lunchtime) regardless of the day, while pink lady apples are bought before noon on Mondays, Tuesdays, Thursdays and Saturdays.**

# Problem 2

```{r p2_data}
data("brfss_smart2010")
```

First, do some data cleaning:

- format the data to use appropriate variable names;
- focus on the “Overall Health” topic
- include only responses from “Excellent” to “Poor”
- organize responses as a factor taking levels ordered from “Poor” to “Excellent”

```{r p2_data_cleaning}
brfss_smart2010 = brfss_smart2010 %>%
  janitor::clean_names() %>%
  filter(topic == "Overall Health" &
           !is.na(response)) %>%
  mutate(response = factor(response, levels = c("Poor", "Fair", "Good", 
                                "Very good", "Excellent")))
```


Using this dataset, do or answer the following (commenting on the results of each):

- In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r p2_locations}
brfss_smart2010 %>%
  filter(year == 2002) %>%
  group_by(locationabbr) %>%
  summarize(unique_locations = length(unique(locationdesc))) %>%
  filter(unique_locations >= 7) %>%
  knitr::kable(caption = "States observed at 7 or more locations in 2002")

brfss_smart2010 %>%
  filter(year == 2010) %>%
  group_by(locationabbr) %>%
  summarize(unique_locations = length(unique(locationdesc))) %>%
  filter(unique_locations >= 7)%>%
  knitr::kable(caption = "States observed at 7 or more locations in 2010")
```
**While only 6 states where observed in seven or more locations in 2002 (Connecticut, Florida, Massachusetts, North Carolina, New Jersey and Pennsylvania), 14 states did the same in 2010 (including the same 6 that completed the feature in 2002).**
    
- Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r p2_spaghetti_plot}
brfss_smart2010 %>% 
  filter(response == "Excellent") %>%
  group_by(year, locationabbr) %>%
  summarize(mean_value = mean(data_value))  %>%
  arrange(desc(year)) %>%
  ggplot() + geom_line(aes(x = year, y = mean_value,
                            group = locationabbr)) +
  labs(y = "Mean values across locations", x= "Year", 
       title = "There is a decrease in value between 2002 and 2005 before stabilizing")
```

**The general trend for all states is that the mean value across locations decreased from 2002 to 2005, but then stabilized from then to 2010.**
    
    
- Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

```{r p2_NY_plot}
brfss_smart2010 %>%
  filter(locationabbr == "NY" & (year == 2006 | year == 2010))  %>%
  mutate(year = factor(year, levels = c("2006", "2010")))  %>%
  ggplot() + geom_density(aes(x = data_value, fill = response), alpha= 0.5) +
  facet_grid( ~ year) + 
  labs(x= "Value", y = "Density", 
       title = "Distribution of values in 2006 and 2010 was very similar", 
       fill = "Response")
```

**In both 2006 and 2010, the lowest values correspond to "Poor" responses, while "Fair" and "Excellent" responses follow. In both years, "Good" and "Very Good" responses follow. In 2010, there are less "Poor" responses than in 2006.**

# Problem 3

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

```{r p3_data_cleaning}
df_dem = read_csv(file = "./data/nhanes_covar.csv",
                  skip = 4) %>% 
  janitor::clean_names() %>%
  drop_na() %>%
  filter(age > 20) %>%
  mutate(sex = case_when(
    sex == 1 ~ "male",
    sex == 2 ~ "female"),
    education = case_when(
      education == 1 ~ "Less than high school",
      education == 2 ~ "High school equivalent",
      education == 3 ~ "More than high school"
    )) %>%
  mutate(sex = factor(sex),
         education = factor(education, levels = c("Less than high school",
                                                  "High school equivalent",
                                                  "More than high school")))

df_acc = read_csv(file = "./data/nhanes_accel.csv") %>% 
  janitor::clean_names() %>%
  pivot_longer(min1:min1440, names_to = "minute", values_to = "mims") %>%
  mutate(minute = as.numeric(str_sub(minute, 4, -1)))

df_mims = left_join(df_dem, df_acc, by = "seqn")
```

Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

```{r p3_sex_education}
df_dem %>%
  group_by(sex, education) %>%
  summarize(count = n()) %>%
  knitr::kable()

df_dem %>%
  ggplot() + geom_bar(aes(x = education, fill = sex)) + 
  labs(y = "Count", x = "Education", 
       title = "Most of the subjects have more than high school education",
       subtitle = "Males and females are evenly distributed in two of the three education levels")
```

**We can see from the bar plot that most of the subjects in the study had more than a high school education (about 50% of the study). Although the distribution of sex in the groups of "less than high school" and "more than high school" education is pretty much even, the study had twelve more males with a "high school equivalent" than females with the same level of education.**

Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r p3_age_activity_plot}
df_mims %>%
  group_by(seqn, education, age, sex) %>%
  summarize(total_activity = sum (mims)) %>%
  ggplot() + geom_point(aes(y = total_activity, x = age, 
                            color = sex), alpha = .5) +
  geom_smooth(aes(y = total_activity, x = age, 
                            color = sex),se = FALSE) +
  facet_grid( ~ education) + 
  labs(x = "Age", y = "Total activity", 
       title = "Male subjects were more active only among those without HS education", color = "Sex")
```

**In this plot, we see that female subjects had higher activity levels regardless of age when the subjects had either a high school education, or higher. This includes a very high difference between female and male subjects with a high school education and at around 40 years of age, and among those 60 years of age and with a level of education greater than high school. This trend is not true for subjects with less than high school, where males seemed to be more active. In general, those with an education greater than high school had constant activity levels, across age values, while the other education levels showed a steeper decline once subjects got to an older age.**


Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r p3_24hr_plot}
df_mims %>%
  ggplot() + geom_point(aes(y = mims, x = minute, 
                            color = sex), alpha = .05, size = 0.5)+
  geom_smooth(aes(y = mims, x = minute, 
                            color = sex),se = FALSE) +
  facet_grid( ~ education) +
  labs(x = "Minute of the day", y= "Activity level (MIMS)",
       title = "Subjects exhibited similar trends regardless of education level or sex",
       color = "Sex")
```

**In the plot, we can see that the subjects exhibited similar activity levels regardless of their level of education or sex: almost no activity at around 4 am, with most activity starting at around 5 am and the peak activity time being the usual 8 am to 8 pm. The biggest difference (albeit small) difference can be seen between female and male subjects with an education level greater than high school, where female subjects remained more active than male subjects throughout the day (on similar trends).**
